# Reproducible Research :Peer Assessment -1


## Introduction 
This document presents the results of peer assessments 1 of course [Reproducible Research](https://www.coursera.org/course/repdata) on coursera. This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

This document presents the results of the Reproducible Research's Peer Assessment 1 in a report using a single R markdown document that can be processed by **knitr** and be transformed into an HTML file.

Through this report you can see that activities on weekdays mostly follow a work related routine, where we find some more intensity activity in little a free time that the employ can made some sport.

An important consideration is the fact of our data presents as a t-student distribution (see both histograms), it means that the impact of imputing missing values with the mean has a good impact on our predictions without a significant distortion in the distribution of the data.

##Preparing R Environment to create the Document
Throughout this report when writing code chunks in the R markdown document,using **echo = TRUE** globally in this document.

```{r}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```
##Loading the Libraries that are required for loading the data
```{r}
library(data.table)
library(ggplot2) # using ggplot for representation
```
##Getting and Cleaning the data for our analysis
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

```{r}
rep_data<-read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric")) # loading the data
rep_data$date <- as.Date(rep_data$date, format = "%Y-%m-%d")
rep_data$interval <- as.factor(rep_data$interval)# cleaning the data into Date and Factor format for ease of use.
```
##What is mean total number of steps taken per day?
1.Calculating total steps per day.
```{r}
total_steps_per_day <- aggregate(steps ~ date, rep_data, sum)
colnames(total_steps_per_day) <- c("date","steps")
head(total_steps_per_day)
```
2.Making a Histogram for total number of steps taken per day.
```{r}
ggplot(total_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Histogram of total number of Steps Taken per Day", 
             x = "Total Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw()
```

3.calculating mean and median of the total number of steps taken per day
```{r}
total_steps_per_day_mean <- mean(total_steps_per_day$steps)
total_steps_per_day_median <- median(total_steps_per_day$steps)
total_steps_per_day_mean # mean
total_steps_per_day_median # median

```

##What is the average daily activity pattern?
Calculating the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called steps_per_interval.
```{r}
steps_per_interval <- aggregate(rep_data$steps, 
                                by = list(interval = rep_data$interval),
                                FUN=mean, na.rm=TRUE)
#convert to integers
##this helps in plotting
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```

Now making the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals:
```{r}

ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="red", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
```
Now, we find the 5-minute interval with the containing the maximum number of steps:

```{r}
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]
max_interval
```

Therefore ,the 835th interval has maximum 206 steps.

#Imputing missing values

#1. Total number of missing values:

The total number of missing values in steps can be calculated using **is.na()** method to check whether the value is missing and then summing the logical vector.


```{r}
missing_vals <- sum(is.na(rep_data$steps))
missing_vals
```
#2. Strategy for filling in all of the missing values in the dataset


To populate missing values, we choose to replace them with the mean value at the same interval across days. In most of the cases the median is a better centrality measure than mean, but in our case the total median is not much far away from total mean, and probably we can make the mean and median meets.

We are creating a function na_fill (data, pervalue) which the data arguement is the rep_data data frame and pervalue arguement is the steps_per_interval data frame.
```{r}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rep_data_fill <- data.frame(  
        steps = na_fill(rep_data, steps_per_interval),  
        date = rep_data$date,  
        interval = rep_data$interval)
str(rep_data_fill)
```
Now let us check for any missing values 
```{r}
sum(is.na(rep_data_fill$steps))
```
#3. A histogram of the total number of steps taken each day
Now let us plot a histogram of the daily total number of steps taken, plotted with a bin interval of 1000 steps, after filling missing values.

```{r}
fill_steps_per_day <- aggregate(steps ~ date, rep_data_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```
#Calculate and report the mean and median total number of steps taken per day.
```{r}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```
#Do these values differ from the estimates from the first part of the assignment?

Yes, these values do differ slightly.

**Before filling the data**

Mean : 10766.189
Median: 10765
**After filling the data**

Mean : 10766.189
Median: 10766.189
We see that the values after filling the data mean and median are equal.


#What is the impact of imputing missing data on the estimates of the total daily number of steps?

As you can see, comparing with the calculations done in the first section of this document, we observe that while the mean value remains unchanged, the median value has shifted and virtual matches to the mean.

Since our data has shown a t-student distribution (see both histograms), it seems that the impact of imputing missing values has increase our peak, but it's not affect negatively our predictions.


#Are there differences in activity patterns between weekdays and weekends?

We do this comparison with the table with filled-in missing values.
1. Augment the table with a column that indicates the day of the week
2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).
3. Tabulate the average steps per interval for each data set.
4. Plot the two data sets side by side for comparison.

```{r}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rep_data_fill)
```

Below you can see the panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:
```{r}

ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="blue") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()

```


We can see at the graph above that activity on the weekday has the greatest peak from all steps intervals. But, we can see too that weekends activities has more peaks over a hundred than weekday. This could be due to the fact that activities on weekdays mostly follow a work related routine, where we find some more intensity activity in little a free time that the employ can made some sport. In the other hand, at weekend we can see better distribution of effort along the time.